---
title: "Analysis of NOAA Weather Events"
Author: "Marc Haserodt"
Date: "February 10, 2017"
output: 
  html_document:
    keep_md: true
---
#####Marc Haserodt
#####February 10, 2017
### 1. Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

### 2. Synopsis

The NOAA Storm Database incorporates storm event data such as date, location, number of victims, and property damage estimates.  This report analyzes which events have the most financial impact and are the most dangerous to people.

The inference of this research is that there is little correlation between the economic impact and human impact.  Heat Events, Tornados, Lightning + T-Storm, Flooding, and Cold Events are the top 5 sources of fatalities, while Flooding, and Hurricane + Wind, combined account for over 80% of all economic impact.

### 3. Process the Data

##### 3-1. Set working directory and download the data from source
```{r}
setwd("C:/Users/marc.haserodt/RepData_PeerAssessment2")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "StormData.csv.bz2")
```

##### 3-2. Load libraries
```{r}
library(ggplot2)
library(grid)
library(gridExtra)
library(plyr)
library(R.utils)
library(RCurl)
library(reshape2)
library(scales)
```
##### 3-3. Read the data file
Read the source .csv file
```{r, cache=TRUE}
  stormCsv <- read.csv("StormData.csv.bz2")
```

##### 3-4. Remove unnecessary columns 
```{r}
  keepColumns <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
  stormTrimmed <- stormCsv[, keepColumns]
  stormTrimmed <- rename(stormTrimmed, c("FATALITIES"="fatalities", "INJURIES"="injuries"))
head(stormTrimmed)
```

##### 3-5. Manipulate BGN_DATE
There is a greater number of observations in later years which could skew result.  Setting cutoff at 80% to capture most observations.
```{r}

  totalObs <- nrow(stormTrimmed)
  trimPercent = 0.80
  trimObs = round(totalObs * trimPercent)
  
  stormTrimmed$year = as.numeric(format(as.Date(stormTrimmed$BGN_DATE, format = "%m/%d/%Y"), "%Y"))
  
  # create dataset with annual count 
  yearRecords <- count(stormTrimmed, "year")
  yearRecords <- yearRecords[order(yearRecords$year, decreasing=TRUE), ]
  yearRecords$runningTotal = cumsum(yearRecords$freq)
  cutOffYear <- min(yearRecords[yearRecords$runningTotal < trimObs, 1])
  
  # trim the dataset further
  stormTrimmed <- stormTrimmed[stormTrimmed$year >= cutOffYear, ]
  endYear <- max(stormTrimmed$year)
  
  # clean stormTrimmed
  stormTrimmed$BGN_DATE <- NULL
  rownames(stormTrimmed) <- NULL
head(stormTrimmed)
```

##### 3-6. Manipulate EVTYPE 
The event types have several categories that are similar and can be grouped together to further clean and reduce the dataset.
```{r, cache=TRUE}

  stormTrimmed$damageSource <- NA
  
  stormTrimmed[grepl("precipitation|rain|hail|drizzle|wet|percip|burst|depression|fog|wall cloud", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Wet Precipitation"
  stormTrimmed[grepl("wind|storm|wnd|hurricane|typhoon", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Hurricane + Wind"
  stormTrimmed[grepl("slide|erosion|slump", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Landslide"
  stormTrimmed[grepl("warmth|warm|heat|dry|hot|drought|thermia|temperature record|record temperature|record high", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Heat Events"
  stormTrimmed[grepl("cold|cool|ice|icy|frost|freeze|snow|winter|wintry|wintery|blizzard|chill|freezing|avalanche|glaze|sleet", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Cold Events"
  stormTrimmed[grepl("flood|surf|blow-out|swells|fld|dam break", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Flooding"
  stormTrimmed[grepl("seas|high water|tide|tsunami|wave|current|marine|drowning", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Drowning"
  stormTrimmed[grepl("dust|saharan", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Dust Storms"  
  stormTrimmed[grepl("tstm|thunderstorm|lightning", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Lightning + T-Storm"
  stormTrimmed[grepl("tornado|spout|funnel|whirlwind", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Tornados"
  stormTrimmed[grepl("fire|smoke|volcanic", 
                         stormTrimmed$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Fire + Volcano"
  
  
  stormTrimmed <- stormTrimmed[complete.cases(stormTrimmed[, "damageSource"]), ]
  stormTrimmed$damageSource <- as.factor(stormTrimmed$damageSource)
  
  
  stormTrimmed$EVTYPE <- NULL

  head(stormTrimmed)

```

##### 3-7. Convert CROPDMG, CROPDMGEXP, PROPDMG, and PROPDMGEXP to absolute values

```{r, cache=TRUE}

  toTenPower <- function(x){
    if(is.numeric(x)) {
      x <- x
    }
    else if(grepl("h", x, ignore.case=TRUE)) {
      x <- 2
    }
    else if(grepl("k", x, ignore.case=TRUE)) {
      x <- 3
    }
    else if(grepl("m", x, ignore.case=TRUE)) {
      x <- 6
    }
    else if(grepl("b", x, ignore.case=TRUE)) {
      x <- 9
    }
    else if(x == "" || x == " "){
      x <- 0
    }
    else{
      x <- NA
    }
    x
  }
   
  
  calcAmount <- function(num, exp){
    pow <- toTenPower(exp)
    if(is.numeric(num)){
      num <- num * (10 ^ pow)
    }
    
    if(!is.numeric(num)){
      num <- 0
    }
    
    num
  }
  

  stormTrimmed$propDamage <- mapply(calcAmount, stormTrimmed$PROPDMG, stormTrimmed$PROPDMGEXP)
  stormTrimmed$cropDamage <- mapply(calcAmount, stormTrimmed$CROPDMG, stormTrimmed$CROPDMGEXP)
  stormTrimmed$damageTotal = stormTrimmed$propDamage + stormTrimmed$cropDamage
  
  
  stormTrimmed$PROPDMG <- NULL
  stormTrimmed$PROPDMGEXP <- NULL
  stormTrimmed$CROPDMG <- NULL
  stormTrimmed$CROPDMGEXP <- NULL
head(stormTrimmed)
```

##### 3-8. Create dataset and variables for plots
```{r}
  
  ecoDamage <- aggregate(formula=cbind(propDamage, cropDamage, damageTotal) ~ damageSource, data=stormTrimmed, FUN=sum, na.rm=TRUE)
  ecoDamage <- ecoDamage[order(ecoDamage$damageTotal, decreasing=TRUE),]
  rownames(ecoDamage) <- NULL
  ecoDamage$damageSource <- factor(ecoDamage$damageSource, levels=rev(ecoDamage$damageSource))
  
  
  ecoDamageMelt <- melt(ecoDamage, id.vars=c("damageSource"), measure.vars=c("propDamage","cropDamage"), variable.name="damageType", value.name="damage")
  levels(ecoDamageMelt$damageType)[levels(ecoDamageMelt$damageType)=="propDamage"] <- "property"
  levels(ecoDamageMelt$damageType)[levels(ecoDamageMelt$damageType)=="cropDamage"] <- "crops"
  
  
  humanDamage <-aggregate(formula=cbind(injuries, fatalities) ~ damageSource, data=stormTrimmed, FUN=sum, na.rm=TRUE) 
  humanDamage <- humanDamage[order(humanDamage$injuries, decreasing=TRUE),]
  rownames(humanDamage) <- NULL
  humanDamage$damageSource <- factor(humanDamage$damageSource, levels=rev(humanDamage$damageSource))
  
  maxInjury <- max(humanDamage$injuries)
  maxInjury <- maxInjury + round(maxInjury * 0.25)
 
  maxFatality <- max(humanDamage$fatalities)
  maxFatality <- maxFatality + round(maxFatality * 0.25)  

```

##### 3-10. Save to RData
```{r}
  save(stormTrimmed, 
       humanDamage, 
       ecoDamageMelt,
       ecoDamage, 
       maxInjury, 
       maxFatality,
       cutOffYear,
       endYear,
       file="StormData.RData")
```  


### 4. Results
##### 4-1. Show the first & last 5 lines of the new data set

```{r}
head(stormTrimmed, n=5L)
tail(stormTrimmed, n=5L)
```

##### 4-2. Examine injuries v fatalities
```{r, fig.width=10, warning=FALSE}
g.mid <- ggplot(data=humanDamage, aes(x=1,y=damageSource)) +
            geom_text(aes(label=damageSource), size=4) +
            ggtitle("") +
            ylab(NULL) +
            scale_x_continuous(expand=c(0,0),limits=c(0.94,1.065)) +
            theme(axis.title=element_blank(),
                  panel.grid=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks.y=element_blank(),
                  panel.background=element_blank(),
                  axis.text.x=element_text(color=NA),
                  axis.ticks.x=element_line(color=NA),
                  plot.margin = unit(c(1,-1,1,-1), "mm"))

# add left chart with injuries
g.injuries <- ggplot(data=humanDamage, aes(x=damageSource, y=injuries)) +
            geom_bar(stat = "identity") + 
            geom_text(aes(label=injuries), size=3, vjust=0.5, hjust=2.0) +
            ggtitle("Injuries") +
            scale_y_reverse(expand=c(0, 0), limits=c(maxInjury,0)) + 
            coord_flip() +
            theme(axis.title.x = element_blank(), 
                  axis.title.y = element_blank(), 
                  axis.text.y = element_blank(), 
                  axis.ticks.y = element_blank(), 
                  plot.margin = unit(c(1,-1,1,0), "mm")) 

# add right chart with fatalities
g.fatalities <- ggplot(data=humanDamage, aes(x=damageSource, y=fatalities)) +
            geom_bar(stat = "identity") + 
            geom_text(aes(label=fatalities), size=3, vjust=0.5, hjust=-1.0) +
            ggtitle("Fatalities") +
            scale_y_continuous(expand=c(0, 0), limits=c(0,maxFatality)) + 
            coord_flip() +
            theme(axis.title.x = element_blank(), 
                  axis.title.y = element_blank(), 
                  axis.text.y = element_blank(), 
                  axis.ticks.y = element_blank(), 
                  plot.margin = unit(c(1,0,1,-1), "mm")) 

# combine charts in one plot
gg.injuries <- ggplot_gtable(ggplot_build(g.injuries))
gg.fatalities <- ggplot_gtable(ggplot_build(g.fatalities))
gg.mid <- ggplot_gtable(ggplot_build(g.mid))

grid.arrange(gg.injuries,gg.mid,gg.fatalities,
             ncol=3,widths=c(4/10,2/10,4/10),
             top=paste("Aggregated human injuries & fatalities for weather events from ",cutOffYear," to ",endYear, sep=""))
```

The data
```{r}
humanDamage
```

##### 4.3. Economic Damage
Crop damage is hardly a factor in comparission to the total economic cost of certain weather events, except for Heat & Drought, where it effects more than 90%
The real interesting ones are Wind & Storm and Flooding & High Surf covering together more than 80% of all economic damage over all the years.
Tornado's,Thunderstorms and Snow & Ice, which have high impact in human costs, hardly matter in economic damages
```{r, fig.width=10}
ggplot(ecoDamageMelt, aes(x=damageSource, y=damage/1000000)) + 
  geom_bar(stat = "identity", aes(fill=damageType)) +
  xlab("Event Type") +
  theme(axis.text.x = element_text(angle = 45, size=8, hjust = 1, vjust = 1)) +
  ylab("Total Damage (millions of USD)") +
  ggtitle(paste("Aggregated property and crop damage for weather events from ",cutOffYear," to ",endYear, sep=""))
  

```

The data
```{r}
ecoDamage
```